---
title: "shinyNHSWorkshopOct18"
author: "Chris Beeley"
date: "8 October 2018"
output: 
  ioslides_presentation:
    css: temp.css
runtime: shiny
---

```{r setup, include=FALSE}

library(DiagrammeR)

knitr::opts_chunk$set(echo = FALSE)

```

## Introduction

- Chris Beeley from Nottinghamshire Healthcare
- Using R for 10 years
- Shiny for 6 (when all this were fields)
- Author of Hands on dashboard development with Shiny
- Author of Web Application Development with R using Shiny, 3rd edition out now!

## Today

- Shiny basics
- Shiny layout functions
- Some neat tricks
- Advanced Shiny

## Minimal example

```{r}

shinyAppDir("minimal/")

```

## Reactive programming

- When the inputs change, the outputs change
- Outputs *take a dependency on* inputs
- The Bad Old Days
- What you lose
- E.g. CRUD is hard
- You can control reactive dependencies (more later)

## Structure

- server.R
- ui.R
- app.R (we don't talk about app.R)

## server.R

```{r eval = FALSE, echo = TRUE}

function(input, output){
  
  output$distPlot <- renderPlot({
    
    # generate bins based on input$bins from ui.R
    x    <- faithful[, 2] 
    bins <- seq(min(x), max(x), length.out = input$bins + 1)
    
    # draw the histogram with the specified number of bins
    hist(x, breaks = bins, col = 'darkgray', border = 'white')
    
  })
}

```

## ui.R

```{r eval = FALSE, echo = TRUE}

fluidPage(
  
  titlePanel("Old Faithful Geyser Data"),
  
  sidebarLayout(
    sidebarPanel(
      sliderInput("bins",
                  "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)
    ),
    
    mainPanel(
      plotOutput("distPlot")
    )
  )
)


```

## Widgets

```{r}

shinyAppDir("widgets")

```

## Plots

```{r eval = FALSE, echo = TRUE}

output$plotDisplay = renderPlot({
  
  theData = data.frame("Date" = seq(input$dateRange[1], 
                                    input$dateRange[2], "days"))
  
  if(input$boxInput){
    
    theData$y = ((1 : nrow(theData)) ^ 2) * runif(nrow(theData))
    
  } else {
    
    theData$y = (1 : nrow(theData)) * runif(nrow(theData))
    
  }
  
  p = ggplot(theData, aes(x = Date, y = y)) + 
    geom_line() + 
    geom_smooth(method = input$pickRadio) +
    ggtitle(input$comment)
  
  # don't print, return!
  
  return(p)
})
```

## Text

```{r eval = FALSE, echo = TRUE}

output$textDisplay = renderText({
  
  if(input$theDate == Sys.Date()) {
    
    storyDate = "today"
  } else if(input$theDate < Sys.Date()){
    
    storyDate = "in the past"
  } else if(input$theDate > Sys.Date()){
    
    storyDate = "in the future"
  }
  
  pasteString = c("This story is written", storyDate,
                  "it concerns", input$comboBox, 
                  ". Basically it goes like this:",
                  input$textArea, ". The story should be told", 
                  input$pickNumber, "times.")
  
  pasteString = rep(pasteString, input$pickNumber)
})

```

## Tables

```{r, eval = FALSE, echo = TRUE}

# The DT package
# Explicitly load and use renderDT and DTOutput

output$tableDisplay = renderDT({
  
  getMat = matrix(c(input$slider, class(input$slider),
                    paste(input$checkGroup, collapse = ','),
                    
                    #... edited
                    
                    input$comboBox, class(input$comboBox),
                    input$comment, class(input$comment)
  ), ncol = 2, byrow = TRUE)
  
  getFrame = data.frame(1 : nrow(getMat), getMat)
  
  names(getFrame) = c("rownames", "Value", "Class")
  
  return(getFrame[1 : input$slider, input$checkGroup])
})

```

## Tabset panels

```{r eval = FALSE, echo = TRUE}

mainPanel(
  tabsetPanel(
    tabPanel("Plot", 
             # ...edited
             plotOutput("plotDisplay"),
             textInput(inputId = "comment",
                       label = "textInput",
                       value = "Graph title")),
    
    tabPanel("Table", sliderInput(inputId = "slider",
                                  label = "Number of rows",
                                  min = 1, max = 9, value = 9),
             dataTableOutput("tableDisplay")),
    
    tabPanel("Text", dateInput(inputId = "theDate",
                               label = "Story date"),
             # ...edited
             textOutput("textDisplay"))
  )
)
```

## Layout functions

- Simple
- flowLayout()
- verticalLayout()
- splitLayout()
- Complete
- sidebarLayout()
- navbarPage()
- navlistPanel()
- Do it yourself
- fluidPage()
- fluidRow()
- column()
- Mix them together

## Example

```{r}

shinyAppDir("navBar/")

```

## Code

```{r eval = FALSE, echo = TRUE}

server = function(input, output) {
  
  output$table = renderTable({
    
    head(iris)
  })
}

ui = navbarPage(
  "Navbar demo",
  tabPanel("Inputs",
           sliderInput("slider", "Slider", 
                       min = 1, max = 100, value = 50),
           textInput("text", "Text")),
  tabPanel("Table", tableOutput("table"))
)

shinyApp(ui, server)

```

## Bootstrap

- fluidRow()
- Columns up to width 12

```{r}

shinyAppDir("commentsApp/")

```

## ui.R

```{r eval = FALSE, echo = TRUE}

fluidPage(
  
  title = "Comments application",
  
  fluidRow(
    column(2, selectInput("directorate", "Directorate:", 
                          unique(testData$Directorate2))),
    
    column(5, dataTableOutput("bestComments")),
    
    column(5, dataTableOutput("improveComments"))
  ),
  
  fluidRow(
    column(5, htmlOutput("bestText"), offset = 2),
    column(5, htmlOutput("improveText"))
  )
)

```

## Reactive expressions

- Reactive expressions are expressions that change when their inputs change
- They enable you to maintain the data fetch/ clean code for several outputs in one place
- Crucially, they *cache* their results, which means that it's run once, for all the outputs, and only changes if its inputs change
- Great for computationally intensive or bandwidth heavy data functions

## Standard Shiny

```{r}

node_df <- create_node_df(n = 5 ,
                          label = c("Input 1", "Input 2", "Output 1", 
                                    "Output 2", "Output 3"),
                          fontsize = 24,
                          fillcolor = c("green", "green", "red", "red", "red"),
                          fontcolor = "black")

edge_df <- data.frame("from" = c(1, 1, 2, 2),
                      "to" = c(3, 4, 4, 5))

i_graph_1 <-
  create_graph()

i_graph_2 <-
  i_graph_1 %>%
  add_nodes_from_table(
    table = node_df,
    label_col = label)

i_graph_3 <-
  i_graph_2 %>%
  add_edges_from_table(
    table = edge_df,
    from_col = from,
    to_col = to,
    from_to_map = id_external)

i_graph_3 %>% render_graph()

```

## With reactive object
```{r}

node_dfR <- create_node_df(n = 6,
                           label = c("Input 1", "Input 2", "Reactive object", "Output 1", 
                                     "Output 2", "Output 3"),
                           fontsize = 24,
                           fillcolor = c("green", "green", "blue", "red", "red", "red"),
                           fontcolor = "black")

edge_dfR <- data.frame("from" = c(1, 2, 3, 3, 3),
                       "to" = c(3, 3, 4, 5, 6))

i_graph_1 <-
  create_graph()

i_graph_2 <-
  i_graph_1 %>%
  add_nodes_from_table(
    table = node_dfR,
    label_col = label)

i_graph_3 <-
  i_graph_2 %>%
  add_edges_from_table(
    table = edge_dfR,
    from_col = from,
    to_col = to,
    from_to_map = id_external)

i_graph_3 %>% render_graph()

```

## Dashboards

```{r}

shinyAppDir("dashboard/")

```

## Structure

```{r, eval = FALSE, echo = TRUE}

library(shinydashboard)

dashboardPage(
  dashboardHeader(),
  dashboardSidebar(),
  dashboardBody()
)

```

## Boxes

```{r, eval = FALSE, echo = TRUE}

body = dashboardBody(
  fluidRow(
    box(
      title = "Table", status = "success", width = 8, solidHeader = TRUE, 
      DT::dataTableOutput("moviePicker")
    ),
    box(
      title = "More info", background = "olive", width = 4, collapsible = TRUE, 
      uiOutput("listMovies")
    )
  ),
  
  fluidRow(
    tabBox(width = 5, title = "Graphs", #...
    ),
    box(
      title = "More graphs", width = 4, #...
    ),
    box(
      width = 3, status = "info", # ...
    )
  )
)

```

## Icons and info boxes

```{r, eval = FALSE, echo = TRUE}

  tabItems(
    tabItem(tabName = "icons", icon = icon("linux"),
            fluidRow(
              infoBox("Average budget ($M)", 25, 
                      icon = icon("money"), color = "green"),
              infoBoxOutput("infoBoxYear"),
              infoBoxOutput("infoBoxGenre")
            ),
            fluidRow(
              valueBox("Average budget ($M)", 25, 
                       icon = icon("money"), color = "green"),
              valueBoxOutput("valueBoxYear"),
              valueBoxOutput("valueBoxGenre")
            )),
    tabItem(tabName = "outputs", icon = icon("graph"),
            ### more output here
    ))

```

## Notifications, messages, and tasks

```{r, eval = FALSE, echo = TRUE}

dropdownMenu(type = "tasks", badgeStatus = "info",
             taskItem(value = 20, color = "blue",
                      "Update database"
             ),
             taskItem(value = 80, color = "red",
                      "Review logins"
             ),
             taskItem(value = 30, color = "green",
                      "Organise Christmas party"
             )
),

# dynamic section

dropdownMenuOutput("notifications"),
dropdownMenuOutput("messages")

```


## Debugging

- Start simple
- Make sure your code works outside of Shiny!
- Use of cat()
- Breakpoints and browser()

## Task- build your own

- With gapminder data
- With anything else

## Sharing

- Gist- runGist("6571951")
- GitHub- runGitHub("gapminder", "chrisbeeley")

## Super whizzy tricks

- Reactive UI
- Populate controls
- Control the whole UI, e.g. passwords or loading screens
- Controlling reactivity with observe and isolate
- Validate() and req()
- JavaScript
- Crosstalk
- Interacting with tables
- Saving and loading data
- Global.R
- Shiny on webpages
- Bookmarks
- Persistence and databases
- The shinyverse
